Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-10-18 manuscript_latex/manuscript.tex
Replaced cross-references to main manuscript and appendix

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in Appendix A.
The location- and scale-invariant Box-Cox log-likelihood function is:

@@

We then consider the empirical probability of type I errors: the
probability of incorrectly classifying a centrally normal distribution
as being non-normal based on the test statistic \(\tau_{\text{ecn}}\).
Under the assumption that the asymmetric generalised normal
distributions are centrally normal after robust transformation, this
produces the relationship shown in Figure
\ref{fig:empirical-central-normality-type-1-error-rate}. Since for
\(\kappa \leq 0.80\) error curves for both transformation methods are
similar, we fix the value of the central portion \(\kappa\) to 0.80. In
the remaining we will use the empirical central normality test statistic
values defined using robust location- and scale-invariant Yeo-Johnson
transformations, as it is more conservative (see Appendix E). This leads
to test statistic values listed in Table
\ref{tab:empirical-central-normality}

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in Appendix D.

@@

Location- and scale-invariant transformations are realised by
simultaneously optimising three parameters, i.e.~transformation
parameter \(\lambda\), shift parameter \(x_0\) and scale parameter
\(s\). We derived the log-likelihood function to facilitate optimisation
using MLE. Alternatively, standardisation of a numeric variable (e.g.,
through subtracting its median value and division by its interquartile
range) prior to conventional power transformations may achieve a similar
effect in reducing sensitivity to the distribution's location and scale.
While this alternative helps prevent these issues -- provided that
normalisation does not lead to negative values for Box-Cox
transformation -- location- and scale-invariant transformations seem to
provide an overall better transformation to normality (Appendix F).

@@

This work has the following limitations. Firstly, we did observe several
numerical stability issues for optimisation criteria other than MLE
(Appendix B). These appear in regions where transformation parameters
would lead to very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed. Secondly,
the empirical central normality test is based on simulations instead of
statistical theory, and relies on a somewhat arbitrary definition of the
central portion of a distribution. Thus, while the test may asses
whether data is sufficiently normally distributed for practical
purposes, it should not be used as a strict test for normality.

@@

Invariance of location- and scale-invariant power transformations was
assessed using the optimisation criteria in section
\ref{appendix-b-optimisation-of-transformation-parameters}. This follows the simulation in the main manuscript, where MLE was
used for optimization. In short, we first randomly drew \(10000\) values
from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed normal distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).
We then computed transformation parameter \(\lambda\) using the original
definitions (equations \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (equations \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution using different
optimisation criteria. To assess location invariance, a positive value
\(d_{\text{shift}}\) was added to each distribution with
\(d_{\text{shift}} \in [1, 10^6]\). Similarly, to assess scale
invariance, each distribution was multiplied by a positive value
\(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

@@

The results are shown in Figure
\ref{fig:empirical-central-normality-test-appendix} and Table
\ref {tab:empirical-central-normality-appendix}. These indicate that the
test behaves similarly for the different datasets. For low type I error
rates, the test statistic proposed in the main manuscript is more
conservative than alternatives based on residuals after Box-Cox
transformations of asymmetric generalised normally distributed features
or on residuals from strictly normally distributed features.

>>>>

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in \ref{appendix-a-log-likelihood-functions-for-location-and-scale-invariant-power-transformation}.
The location- and scale-invariant Box-Cox log-likelihood function is:

@@

We then consider the empirical probability of type I errors: the
probability of incorrectly classifying a centrally normal distribution
as being non-normal based on the test statistic \(\tau_{\text{ecn}}\).
Under the assumption that the asymmetric generalised normal
distributions are centrally normal after robust transformation, this
produces the relationship shown in Figure
\ref{fig:empirical-central-normality-type-1-error-rate}. Since for
\(\kappa \leq 0.80\) error curves for both transformation methods are
similar, we fix the value of the central portion \(\kappa\) to 0.80. In
the remaining we will use the empirical central normality test statistic
values defined using robust location- and scale-invariant Yeo-Johnson
transformations, as it is more conservative (see \ref{appendix-e-empirical-central-normality-test}). This leads
to test statistic values listed in Table
\ref{tab:empirical-central-normality}

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in 
\ref{appendix-d-experimental-results-using-location--and-scale-invariant-box-cox-transformation}.

@@

Location- and scale-invariant transformations are realised by
simultaneously optimising three parameters, i.e.~transformation
parameter \(\lambda\), shift parameter \(x_0\) and scale parameter
\(s\). We derived the log-likelihood function to facilitate optimisation
using MLE. Alternatively, standardisation of a numeric variable (e.g.,
through subtracting its median value and division by its interquartile
range) prior to conventional power transformations may achieve a similar
effect in reducing sensitivity to the distribution's location and scale.
While this alternative helps prevent these issues -- provided that
normalisation does not lead to negative values for Box-Cox
transformation -- location- and scale-invariant transformations seem to
provide an overall better transformation to normality 
(\ref{appendix-f-normalisation-before-transformation}).

@@

This work has the following limitations. Firstly, we did observe several
numerical stability issues for optimisation criteria other than MLE
(\ref{appendix-b-optimisation-of-transformation-parameters}).
These appear in regions where transformation parameters
would lead to very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed. Secondly,
the empirical central normality test is based on simulations instead of
statistical theory, and relies on a somewhat arbitrary definition of the
central portion of a distribution. Thus, while the test may asses
whether data is sufficiently normally distributed for practical
purposes, it should not be used as a strict test for normality.

@@

Invariance of location- and scale-invariant power transformations was
assessed using the optimisation criteria in 
\ref{appendix-b-optimisation-of-transformation-parameters}. This follows the simulation in section \ref{invariance-to-location-and-scale}, where MLE was
used for optimization. In short, we first randomly drew \(10000\) values
from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed normal distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).
We then computed transformation parameter \(\lambda\) using the original
definitions (equations \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (equations \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution using different
optimisation criteria. To assess location invariance, a positive value
\(d_{\text{shift}}\) was added to each distribution with
\(d_{\text{shift}} \in [1, 10^6]\). Similarly, to assess scale
invariance, each distribution was multiplied by a positive value
\(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

@@

The results are shown in Figure
\ref{fig:empirical-central-normality-test-appendix} and Table
\ref {tab:empirical-central-normality-appendix}. These indicate that the
test behaves similarly for the different datasets. For low type I error
rates, the test statistic proposed in section \ref{empirical-central-normality-test-1} is more
conservative than alternatives based on residuals after Box-Cox
transformations of asymmetric generalised normally distributed features
or on residuals from strictly normally distributed features.
